Multiple sequence alignment

### 2025-04-04 episodes/msa.Rmd
Corrected a div error

```bash
interactive -A uppmax2024-2-10 -M snowy -t 4:00:00
cd /proj/g2020004/nobackup/3MK013/<username>
mkdir phylogenetics
cd phylogenetics
```

@@

@@

@@

Use the `module` command to load `bioinfo-tools` and `MAFFT`.

@@

```output
...
Strategy:
 FFT-NS-2 (Fast but rough)
 Progressive method (guide trees were built 2 times.)

@@

mafft-fftns rpoB_renamed.fasta > rpoB.fftns.aln  1.02s user 0.52s system 104% cpu 1.471 total
```

The last line is the output of the `time` command. It took 1.47 seconds to complete this time.

@@

Use the E-INS-i algorithm from mafft to align the renamed sequences. Also, record the time it takes for `mafft` to complete the task.

@@

```output
...
Strategy:
 E-INS-i (Suitable for sequences with long unalignable regions, very slow)
 Iterative refinement method (<16) with LOCAL pairwise alignment with generalized affine gap costs (Altschul 1998)

@@

mafft-einsi rpoB_renamed.fasta > rpoB.einsi.aln  6.84s user 0.65s system 99% cpu 7.542 total
```

It now took 7.54 seconds to complete this time, i.e. about 5 times slower than with the progressive algorithm. It doesn't make a big difference now, but with hundreds of sequences it will make one.

@@

@@

>>>>

```bash
interactive -A uppmax2025-3-4 -M snowy -t 4:00:00
cd /proj/g2020004/nobackup/3MK013/<username>
mkdir phylogenetics
cd phylogenetics
```

@@

Use the `../../` to see what's two folders up, and then `data/rpoB/rpoB.fasta`

@@

<!---

--->

@@

Use the `module` command to load `bioinfo-tools` and `MAFFT`. Use `time` to record the time.

The help obtained through `mafft -h` is not very informative about algorithms, so check the [mafft webpage](https://mafft.cbrc.jp/alignment/software/algorithms/algorithms.html).

`mafft` actually has one executable program for each algorithm, all starting with `mafft-`. A way to display them all is to type that and push the Tab key twice to see all possibilities.

@@

```bash
module load <general module> <mafft module>
time mafft-<algorithm> <fasta_file> > rpoB.fftns.aln
```

```output
...
[...]
Strategy:
 FFT-NS-2 (Fast but rough)
 Progressive method (guide trees were built 2 times.)

@@

real	0m1.125s
user	0m0.818s
sys	0m0.181s
```

The last line is the output of the `time` command. It took 1.125 seconds to complete this time.

@@

Use the superior E-INS-i algorithm from mafft to align the renamed sequences. Also, record the time it takes for `mafft` to complete the task.

@@

```bash
time mafft-<better algo> <fasta file> > rpoB.einsi.aln
```

```output
[...]
Strategy:
 E-INS-i (Suitable for sequences with long unalignable regions, very slow)
 Iterative refinement method (<16) with LOCAL pairwise alignment with generalized affine gap costs (Altschul 1998)

@@

real	0m7.367s
user	0m7.022s
sys	0m0.244s
```

It now took 7.36 seconds to complete this time, i.e. 6 times more than with the progressive algorithm. It doesn't make a big difference now, but with hundreds of sequences it will make one.

@@

Use `scp` to copy files from Uppmax to your computer. `scp` allows wildcards, but you probably need to escape the `*`.

```bash
scp <username>@rackham.uppmax.uu.se:/<absolute path to phylogenetics folder>/rpoB.\*.aln <localfolder>/
```

@@

```bash
scp lionel@rackham.uppmax.uu.se:/proj/g2020004/nobackup/3MK013/<user>/phylogenetics/rpoB.\*.aln .
```
